Ecological Limits to Forecasting: Estimating Uncertainty in Critical Transitions with Deep Learning

### 2022-05-14 manuscript/manuscript.Rmd
Small edits

Forecasting plays an important and rapidly growing role in both testing our fundamental understanding of ecological processes, and informing ecological applications and conservation decision-making [@dietze_iterative_2018; @schindler_portfolio_2015].
Meanwhile, recent advances in machine learning have rapidly improved the prevalence and accuracy of short term forecasts in many fields [@kao_exploring_2020; @lyu_lstm_2020; @du_multivariate_2020]. 
Will these emerging methods improve the capacity for forecasts in ecological systems as well?
Ecological dynamics are notoriously complex, with uncertainty and non-linearity playing critical roles [@Boettiger2018; @hallett_why_2004; @Oveskienen2010].
These challenges are nowhere more evident then in _critical transitions_, sudden shifts in the states or patterns of ecosystem dynamics that are more important and more difficult to predict than gradual changes.
Here, we examine several of the best-known examples of critical transitions in ecological systems.
We evaluate the ability of some of the most promising machine learning methods for probabilistic forecasts relative to traditional statistical and mechanistic approaches applied to several classic models in ecology.

@@

The Nicholson-Bailey model describes a predator-prey dynamic for the relationship of a host species and an obligate parsatiod, originally used to model the population dynamics of blowflies (_Lucilia cuprina_) [@nicholson_balance_1935; @nicholson_outline_1954; @nicholson_compensatory_1954].
We consider the form which includes density dependence in the host species, and allow for environmental stochasticity,

@@

Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at linear rate, which drives a Hopf bifurcation as $K$ becomes sufficiently large. 
In a Hopf bifurcation, a stable node starts an oscillatory pattern which continues to grow in amplitude as the bifurcation parameter continues to increase.
This example illustrates one of the many kinds of challenges which nonlinear phenomena pose to forecasting: the "historical" data prior to the bifurcation never exhibit the cyclical dynamics of growing amplitude that will emerge after the bifurcation occurs.
If we had used a purely deterministic model, the dynamics would be constrained to a single stable point, corresponding to a slowly changing steady-state population size of host and parasitoid populations.
However, stochasticity in this case acts as a source of some additional information about the dynamics, as the noise excites quasi-cycles which are visible in the irregular oscillations that appear significantly prior to the emergence of true limit cycles which follow the bifurcation [@boettiger_noise_2018].
Examples use the following parameters: $H_0 = 9$, $P_0 = 1$, $r = 0.75$, $c = 0.1$, $K_0 = 14$, $\delta = 0.08$, $\sigma_H = 0.02$, $\sigma_P = 0.02$.

@@

We examine forecast skill for each of the six forecasting methods (MCMC, ARIMA, block-RNN, GRU, LSTM, and transformer) in each of our three scenarios (Hopf bifurcation, saddle-node bifurcation, stochastic transition).
In addition to these cases, we also consider an "ensemble model", generated by drawing from the distribution of all models except the MCMC model. 
Such ensemble techniques can better reflect uncertainty than relying on any single method [@gneiting_weather_2005].
For simplicity, we consider the unweighted case, where each model is represented equally in the ensemble.
Using model-based simulations allows us to examine performance against a multiple (n=100) replicates of the "true" process, which further helps identify differences that may occur solely due to chance. 
By taking the true model structure as given, MCMC methods can be used to determine a theoretical limit of forecasting skill.
Note that in both bifurcation scenarios, future dynamics will see states never previously observed in the historical data that was used to train each of the methods (i.e. very small population sizes).  
This no-analog aspect of forecasting bifurcation dynamics means that even with many sample points in the training data _and_ perfect knowledge of the true model structure, posterior distributions of parameter values are still influenced by the choice of priors.

>>>>

Forecasting plays an important and rapidly growing role in both testing our fundamental understanding of ecological processes, and informing ecological applications and conservation decision-making [@dietze_iterative_2018; @schindler_portfolio_2015].
Meanwhile, recent advances in machine learning have rapidly improved the prevalence and accuracy of short term forecasts in many fields [@kao_exploring_2020; @lyu_lstm_2020; @du_multivariate_2020]. 
Will these emerging methods improve the capacity for forecasts in ecological systems as well?
Ecological dynamics are notoriously complex, with uncertainty and non-linearity playing critical roles [@Boettiger2018; @hallett_why_2004; @Oveskienen2010].
These challenges are nowhere more evident then in _critical transitions_, sudden shifts in the states or patterns of ecosystem dynamics that are more important and more difficult to predict than gradual changes.
Here, we examine several of the best-known examples of critical transitions in ecological systems.
We evaluate the most promising machine learning methods for probabilistic forecasts relative to traditional statistical and mechanistic approaches applied to several classic models in ecology.

@@

The Nicholson-Bailey model describes a predator-prey dynamic for the relationship of a host species and an obligate parasitoid, originally used to model the population dynamics of blowflies (_Lucilia cuprina_) [@nicholson_balance_1935; @nicholson_outline_1954; @nicholson_compensatory_1954].
We consider the form which includes density dependence in the host species, and allow for environmental stochasticity,

@@

Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at linear rate, which drives a Hopf bifurcation as $K$ becomes sufficiently large. 
In a Hopf bifurcation, a stable node starts an oscillatory pattern which grows in amplitude as the bifurcation parameter continues to increase.
This example illustrates one of the many kinds of challenges which nonlinear phenomena pose to forecasting: the "historical" data prior to the bifurcation never exhibit the cyclical dynamics of growing amplitude that will emerge after the bifurcation occurs.
If we had used a purely deterministic model, the dynamics would be constrained to a single stable point, corresponding to a slowly changing steady-state population size of host and parasitoid populations.
However, stochasticity in this case acts as a source of some additional information about the dynamics, as the noise excites quasi-cycles which are visible in the irregular oscillations that appear significantly prior to the emergence of true limit cycles which follow the bifurcation [@boettiger_noise_2018].
Examples use the following parameters: $H_0 = 9$, $P_0 = 1$, $r = 0.75$, $c = 0.1$, $K_0 = 14$, $\delta = 0.08$, $\sigma_H = 0.02$, $\sigma_P = 0.02$.

@@

We examine forecast skill for each of the six forecasting methods (MCMC, ARIMA, block-RNN, GRU, LSTM, and transformer) in each of our three scenarios (Hopf bifurcation, saddle-node bifurcation, and stochastic transition).
In addition to these cases, we also consider an "ensemble model", generated by drawing from the distribution of all models except the MCMC model. 
Such ensemble techniques can better reflect uncertainty than relying on any single method [@gneiting_weather_2005].
For simplicity, we consider the unweighted case, where each model is represented equally in the ensemble.
Using model-based simulations allows us to examine performance against multiple (n=100) replicates of the "true" process, which further helps identify differences that may occur solely due to chance. 
By taking the true model structure as given, MCMC methods can be used to determine a theoretical limit of forecasting skill.
Note that in both bifurcation scenarios, future dynamics will see states never previously observed in the historical data that was used to train each of the methods (i.e. very small population sizes).  
This no-analog aspect of forecasting bifurcation dynamics means that even with many sample points in the training data _and_ perfect knowledge of the true model structure, posterior distributions of parameter values are still influenced by the choice of priors.
